---
title: "RRPA2"
output: html_document
---
## Synopsis
The purpose of this report is to investigate the types of event are most harmful to the population health and economy. Based on the analysis of the data, it is observed that the top aggregate impact to population health and property to be Tornados and floods have caused the most significant economic damage.

## Data Processing
# Environment Initialization
```{r}
library(ggplot2)
library(xtable)
setwd("C:/Users/Xiaoyu Liu/Dropbox/Onlinecourse/DataAnalysisonline/reproductive")
raw.data.file.name <- "storm-raw-data.csv.bz2"

if (!file.exists(raw.data.file.name)) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",raw.data.file.name, method = "wget", extra = "--no-check-certificate")
}
```

# Load the data

The following fields were used for this analysis:

Field  Description
BGN_DATE	Date the storm event began
EVTYPE	Type of storm event (not abbreviated)
FATALITIES	The number of deaths directly related to the weather event
INJURIES	The number of injuries directly related to the weather event
PROPDMG	The estimated amount of damage to property incurred by the weather event (whole numbers and hundredths)
PROPDMGEXP	A multiplier where "H" denotes hundreds, 'K" denotes thousands, "M" denotes millions, and "B" denotes billions
CROPDMG	The estimated amount of damage to crops incurred by the weather event (whole numbers and hundredths)
CROPDMGEXP	A multiplier where "H" denotes hundreds, 'K" denotes thousands, "M" denotes millions, and "B" denotes billions

```{r}
data.file.name <- "storm-data.csv"

if (!file.exists(data.file.name)) {
    raw.data <- read.csv(bzfile(raw.data.file.name), header = TRUE, stringsAsFactors = FALSE)
    raw.data$BGN_DATE <- as.Date(raw.data$BGN_DATE, format = "%m/%d/%Y %H:%M:%S")
    # Horizontal subset
    data <- raw.data[raw.data$BGN_DATE >= as.Date("1993-01-01"), ]
    data <- data[data$INJURIES > 0 | data$FATALITIES > 0 | data$PROPDMG > 0 | data$CROPDMG > 0, ]
    # Vertical subset - limit the data to the required fields.
    data <- data[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
    # Transform PROPDMG multiplier
    data$PROPDMGEXP[is.na(data$PROPDMGEXP)] <- 0
    data$PROPDMGEXP[data$PROPDMGEXP == ""] <- 1
    data$PROPDMGEXP[grep("[-+?]", data$PROPDMGEXP)] <- 1
    data$PROPDMGEXP[grep("[Hh]", data$PROPDMGEXP)] <- 100
    data$PROPDMGEXP[grep("[Kk]", data$PROPDMGEXP)] <- 1000
    data$PROPDMGEXP[grep("[Mm]", data$PROPDMGEXP)] <- 1e+06
    data$PROPDMGEXP[data$PROPDMGEXP == "B"] <- 1e+09
    data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
    # Transform CROPDMG multiplier
    data$CROPDMGEXP[is.na(data$CROPDMGEXP)] <- 0
    data$CROPDMGEXP[data$CROPDMGEXP == ""] <- 1
    data$CROPDMGEXP[grep("[-+?]", data$CROPDMGEXP)] <- 1
    data$CROPDMGEXP[grep("[Hh]", data$CROPDMGEXP)] <- 100
    data$CROPDMGEXP[grep("[Kk]", data$CROPDMGEXP)] <- 1000
    data$CROPDMGEXP[grep("[Mm]", data$CROPDMGEXP)] <- 1e+06
    data$CROPDMGEXP[data$CROPDMGEXP == "B"] <- 1e+09
    data$CROPDMGEXP <- as.numeric(data$CROPDMGEXP)
    # Adjust property and crop damages
    data$PROPDMG <- data$PROPDMG * data$PROPDMGEXP
    data$CROPDMG <- data$CROPDMG * data$CROPDMGEXP

    # Clean Event Type
    data$EVTYPE <- toupper(data$EVTYPE)
    # Trim multiple spaces and leading, trailing spaces
    trim <- function(x) gsub("[[:space:]]{2,}", " ", gsub("(^[[:space:]]+|[[:space:]]+$)", "", x))
    data$EVTYPE <- trim(data$EVTYPE)
    write.csv(data, file = data.file.name, row.names = FALSE)
} else {
    data <- read.csv(data.file.name, header = TRUE, stringsAsFactors = FALSE)
}
```
## Results
# Question 1. Across the United States, which types of events (as indicated in the [EVTYPE] variable) are most harmful with respect to human health?
```{r}
population.health.data <- data[, c("EVTYPE", "INJURIES", "FATALITIES")]
population.health.data <- aggregate.data.frame(x = population.health.data[, c(2, 3)], by = list(population.health.data$EVTYPE), FUN = sum)
colnames(population.health.data)[1] <- "EVTYPE"
population.health.data$TOTAL <- population.health.data$FATALITIES + population.health.data$INJURIES
population.health.data <- population.health.data[order(population.health.data$TOTAL, decreasing = TRUE), ]
row.names(population.health.data) <- population.health.data$EVTYPE
population.health.data <- population.health.data[population.health.data$TOTAL != 0, ]

```
```{r}
population.health.table <- xtable(population.health.data[1:10, c("INJURIES", "FATALITIES")], caption = "Table 1. Event types that accounted for the highest cumulative injuries and fatalities")
names(population.health.table) <- c("Injuries", "Fatalities")
print(population.health.table, type = "html")
```
The following plot demonstrates that tornados are most the most harmful weather event to population health.

```{r}
population.health.plot <- barplot(height = t(as.matrix(population.health.data[1:10,  c("INJURIES", "FATALITIES")])), main = bquote(atop("Figure 1. Top 10 severe weather events caused injuries and fatalities","U.S. 1993 - 2011")), legend = c("Injuries", "Fatalities"), axes = FALSE,axisnames = FALSE, col = c("yellow", "red"))

text(population.health.plot, par("usr")[3], labels = row.names(population.health.data)[1:10],srt = 45, adj = 1, cex = 0.6, xpd = TRUE)
axis(2)
```

#Question 2. Across the United States, which types of events (as indicated in the [EVTYPE] variable) have the greatest economic consequences?
```{r}
economy.data <- data[, c("EVTYPE", "PROPDMG", "CROPDMG")]
economy.data <- aggregate.data.frame(x = economy.data[, c(2, 3)], by = list(economy.data$EVTYPE),FUN = sum)
colnames(economy.data)[1] <- "EVTYPE"
economy.data$TOTAL <- economy.data$PROPDMG + economy.data$CROPDMG
economy.data <- economy.data[order(economy.data$TOTAL, decreasing = TRUE), ]
row.names(economy.data) <- economy.data$EVTYPE
economy.data <- economy.data[economy.data$TOTAL != 0, ]
```
```{r}
economy.table <- xtable(economy.data[1:10, c("PROPDMG", "CROPDMG")], caption = "Table 2. Event types that accounted for the highest property damage and crop damage")
names(economy.table) <- c("Property Damage", "Crop Damage")
print(economy.table, type = "html")
```
The following plot demonstrates that floods have the greatest economic damages impact on property and crops.
```{r}
economy.plot <- barplot(height = t(as.matrix(economy.data[1:10, c("PROPDMG", "CROPDMG")])), main = bquote(atop("Figure 2. Top 10 weather events caused property damage", "U.S. 1993 - 2011")), legend = c("Properties", "Crops"), axes = FALSE, axisnames = FALSE,col = c("brown", "green"))

text(economy.plot, par("usr")[3], labels = row.names(economy.data)[1:10], srt = 45, adj = 1, cex = 0.6, xpd = TRUE)

axis(2, at = c(0, economy.data[1, "TOTAL"]), labels = c("0.0", format(economy.data[1,"TOTAL"], digits = 2)))
```

## Conclusions
Tornados have the greatest impact on the population health. Flood has the largest impact on the property and second largest impact on crops. Drought has the largest impact on the crops. Overall, flood brings the largest economic consequence.

